---
title: "Exploratory Data Analysis"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

```{r imports, include=F, warning=F, message=F}
library(dplyr)
library(plotly)
library(leaflet)
library(DT)
library(htmltools)
```


```{r import data, include=F}
df = read.csv("data/merged_data.csv", header = TRUE)
coord_df = read.csv("data/stations_data.csv", header = TRUE)
```

```{r delay frequency by station, include=F}
df_map <- df %>%
  inner_join(coord_df, by = c("station" = "station")) %>%
  group_by(station, latitude, longitude, service_population) %>%
  summarize(delay_count = n(), .groups = "drop")

delay_map = leaflet(df_map) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~longitude, lat = ~latitude,
    radius = ~sqrt(delay_count),  # scale size by count
    color = "darkred",
    stroke = FALSE, fillOpacity = 0.7,
    popup = ~paste0("<b>", station, "</b><br>Service Population: ", service_population, "</b><br>Delays: ", delay_count)
  ) %>%
  addControl(
    html = tags$div(
      tags$h4("Subway Delay Frequency by Station")
    ),
    position = "topright"
  )

```


```{r service pop and delay metrics, echo=FALSE}
station_line_stats <- df %>%
  group_by(station, line, service_population) %>%
  summarize(
    delay_count = n(),
    avg_duration = mean(min_delay, na.rm = TRUE),
    .groups = "drop"
  )

scatter_plot = plot_ly(
  data = station_line_stats,
  x = ~service_population,
  y = ~avg_duration,
  type = "scatter",
  mode = "markers",
  color = ~line,
  marker = list(size = 10, opacity = 0.8),
  text = ~paste0(
    "<b>", station, " (", line, ")</b><br>",
    "Service Pop: ", service_population, "<br>",
    "Delays: ", delay_count, "<br>",
    "Avg Duration: ", round(avg_duration, 1), " min"
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Average Delay vs Service Population by Station-Line",
    xaxis = list(title = "Service Population"),
    yaxis = list(title = "Average Delay Duration (min)")
  )
```


```{r delay frequency by day and time, include=F}
heat_df <- df %>%
  group_by(day, hour) %>%
  summarize(delay_count = n(), .groups = "drop")

# Ensure day order (optional)
heat_df$day <- factor(heat_df$day, 
                      levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

heat_plot = plot_ly(heat_df,
        x = ~hour,
        y = ~day,
        z = ~delay_count,
        type = "heatmap",
        colors = "Reds") %>%
  layout(title = "Frequency of Subway Delays by Hour and Day of Week",
         xaxis = list(title = "Hour of Day"),
         yaxis = list(title = "Day of Week"))
```


This page shows some interactive plots created for exploratory purposes. Hover over the plots to see more information. You can find a written EDA report [here](https://github.com/omina26/TTC-Subway-Delay-Analysis/blob/main/EDA/report.pdf).

### Figure A. Delay Map


```{r echo=FALSE, message=FALSE, fig.cap="<span style='font-size: 90%;'><strong>Figure A</strong>. The frequency of TTC subway delays at each station in 2024. Circle size reflects the number of delays, and the popup includes the service population.</span>"}
delay_map
```

According to Figure A, Bloor-Yonge Station had the highest frequency of subway delays across all stations. It also appears that terminal stations had more delays compared to non-terminal stations.


### Figure B. Delay vs Service Population


```{r echo=FALSE, message=FALSE, fig.cap="<span style='font-size: 90%;'><strong>Figure B</strong>. Scatterplot of the average TTC Subway delay duration for each subway and line combination by the service population in 2024. Each point shows a station-line combination, with service population in the x-axis and average delay duration in the y-axis.</span>"}
scatter_plot
```

According to Figure B, there doesn't appear to be a correlation between service population and average delay duration. On average, average delay duration is between 5 and 10 minutes. The stations with average delay duration greater than 10 minutes have one or two delays. Hence, these were probably caused by an unpredictable emergency situation.


### Figure C. Heatmap: Delays by Day and Hour


```{r echo=FALSE, message=FALSE, fig.cap="<span style='font-size: 90%;'><strong>Figure C</strong>. TTC Subway delay frequencys across days of the week and times of the day in 2024. Darker colors indicate more frequent delays during specific hours and days.</span>"}
heat_plot
```

According to Figure C, delays are most frequent on SUndays at 8 AM. Further, morning and afternoon rush hours are more prone to delays during the work week.





